---
title: "Test - sum NB"
output: html_notebook
---


```{r setup}
devtools::load_all()
library(here)
library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
library(tidyverse)
library(cowplot)

```

```{r}
model_sum_nb <- stan_model(here("stan","test_sum_nb.stan"))
```

Generating data

```{r}
generator <- function(G, N, method = "saddlepoint", prior_mean_mus = 2, prior_sd_mus = 1) {
  if(method == "saddlepoint") {
    method_id = 0
  } else if (method == "moments") {
    method_id = 1
  } else {
    stop("Invalid method")
  }
  
  function() {
    all_mus <- rlnorm(G + 1, prior_mean_mus, prior_sd_mus)
    all_mus[G + 1] <- rlnorm(1, 5, 3)
    all_phis <- 1 / sqrt(abs(rnorm(G + 1)))
    sums <- array(-1, N)
    repeats <- 0
    for(n in 1:N) {
      repeat {
        sums[n] <- sum(rnbinom(G + 1, mu = all_mus, size = all_phis))
        if(sums[n] > 0) {
          break;
        } 
        repeats <- repeats + 1
      }
    }
    if(repeats > 0) {
      cat(repeats, " repeats\n")
    }
    list(
      observed = list(
        N = N,
        sums = sums,
        G = G,
        method = method_id,
        mus = array(all_mus[1:G], G),
        phis = array(all_phis[1:G], G)
        #phis = all_phis
      ),
      true = list(
        extra_mu = all_mus[G+1],
        extra_phi = all_phis[G+1]
      )
    )
  }
}
```

Testing a single version to see it doesn't break terribly.

```{r}
data <- generator(G = 5, N = 5, "moments")()
fit <- sampling(model_sum_nb, data$observed)
evaluate_all_params(rstan::extract(fit), data$true)
```

# Sum of two NBs

Here we test a sum of two NBs - one is small and has known parameters, the other has unknown parameters and possibly larger mean (the prior mean is larger).

Calibration and accuracy with saddlepoint approximation

```{r}
sbc_res_saddlepoint_small <- sbc(model_sum_nb, generator(G = 1, N = 10, "saddlepoint"), N_steps = 100, control = list(adapt_delta = 0.95))
plot_sbc_params(sbc_res_saddlepoint_small$params, x_axis_trans = "log10", y_axis_trans = "log10")
summarise_sbc_diagnostics(sbc_res_saddlepoint_small)

```

Median time - saddlepoint small, individual: 52.7035

Calibration and accuracy with moments approximation

```{r}
sbc_res_moments_small <- sbc(model_sum_nb, generator(G = 1, N = 10, "moments"), N_steps = 100, control = list(adapt_delta = 0.95))
plot_sbc_params(sbc_res_moments_small$params, x_axis_trans = "log10", y_axis_trans = "log10")
summarise_sbc_diagnostics(sbc_res_moments_small)

```

# Sum of 21 NBs

20 NBs are known (low means) and one NB is unknown (a prior large mean)

Saddlepoint:

```{r}
sbc_res_saddlepoint_large <- sbc(model_sum_nb, generator(G = 20, N = 20, "saddlepoint"), N_steps = 100, control = list(adapt_delta = 0.95))
plot_sbc_params(sbc_res_saddlepoint_large$params, x_axis_trans = "log10", y_axis_trans = "log10")
summarise_sbc_diagnostics(sbc_res_saddlepoint_large)

```

Median time - saddlepoint large, individual: 259.2945

Moments:

```{r}
sbc_res_moments_large <- sbc(model_sum_nb, generator(G = 20, N = 20, "moments"), N_steps = 100, control = list(adapt_delta = 0.95))
plot_sbc_params(sbc_res_moments_large$params, x_axis_trans = "log10", y_axis_trans = "log10", plot_stat = "mean")
summarise_sbc_diagnostics(sbc_res_moments_large)

```

# 10 large NBs

9 known NBs, 1 unknown, all large ($log(\mu) \sim N(5,3)$). Here, the estimates start to be dominated by the prior.

Saddlepoint:
```{r}
sbc_res_saddlepoint_10_large <- sbc(model_sum_nb, generator(G = 9, N = 10, "saddlepoint", prior_mean_mus = 5, prior_sd_mus = 3), N_steps = 100, control = list(adapt_delta = 0.95))
plot_sbc_params(sbc_res_saddlepoint_10_large$params, x_axis_trans = "log10", y_axis_trans = "log10")
summarise_sbc_diagnostics(sbc_res_saddlepoint_10_large)
```


Moments:
```{r}
sbc_res_moments_10_large <- sbc(model_sum_nb, generator(G = 9, N = 10, "moments", prior_mean_mus = 5, prior_sd_mus = 3), N_steps = 100, control = list(adapt_delta = 0.95))
plot_sbc_params(sbc_res_moments_10_large$params, x_axis_trans = "log10", y_axis_trans = "log10")
summarise_sbc_diagnostics(sbc_res_moments_10_large)
```
