---
title: "Testing Waring"
output: html_notebook
---

```{r setup}
library(rstan)
library(gamlss.dist)
library(tidyverse)
library(cowplot)
library(here)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
devtools::load_all()
```


Testing different parametrizations

Direct parametrization.

```{r}
waring_lpdf <- function(y, mu, k, rho) {
  a <- mu * (rho - 1) / k
  lgamma(a + rho) + lgamma(k + rho) - lgamma(rho) - lgamma(a + k + rho) + lgamma(a + y) - lgamma(a) + lgamma(k + y) - lgamma(k) - lgamma(a + k + rho + y) + lgamma(a + k + rho) - lgamma(y + 1)
}


tibble(mu = 50) %>% 
  crossing(tibble(y = seq(from = 0, to = 1000, by = 1))) %>% 
  crossing(tibble(k = c(0.1, 0.5, 1, 3, 5, 10)))%>% 
  crossing(tibble(rho = seq(1.1, 10, length.out = 5)))%>% 
  mutate(ld = waring_lpdf(y, mu, k, rho))  %>%
    ggplot(aes(x = y, y = ld, color = rho, group = rho)) + geom_line() + facet_wrap(~k, ncol = 3, labeller = "label_both")

```

## Decomposing variance (looks best so far)

The main point is that for $Y \sim Waring(\mu, \rho, k)$ you can decompose the variance as:

$$
Var(Y) = \mu + \tau\mu + \frac{\mu^2}{\phi} \\
\tau = \frac{k + 1}{\rho - 2} \\
\phi = \frac{k ( \rho - 2 )} {k + \rho - 1}
$$ 


So as $\tau \rightarrow 0$ the distribution approaches $NB(\mu, \phi)$. Eventually the $\tau$ parameter controls the shape of the tail in a neat way:


```{r}
main_data <- tibble(mu = 50) %>% 
  crossing(tibble(y = seq(from = 0, to = 1000, by = 1))) %>% 
  crossing(tibble(phi = c(1, 5, 10)))%>% 
  crossing(tibble(tau = c(0.001, 0.1, 1, 10))) %>% 
  mutate(
    k = (tau + 1)*phi,
    rho = (phi + 1) / tau + phi + 2,
    ld = waring_lpdf(y, mu = mu, k = k , rho = rho))

nb_data <- main_data %>% filter(y %% 100 == 0) %>%
  mutate(
         nb_dens = dnbinom(y, mu = mu, size = phi, log = TRUE)
         )

main_data %>%
  ggplot(aes(x = y, y = ld, color = as.factor(phi), group = phi)) + 
      geom_point(data = nb_data, aes(y = nb_dens, shape = as.factor(phi))) +
      geom_line() + 
      facet_wrap(~tau, ncol = 2, labeller = "label_both", scales = "free_y")
```


The lines show the Waring lpdf, the points are the NB lpdf with the same $\mu$ and $\phi$.

The reverse transformation is:

https://www.wolframalpha.com/input/?i=solve+for+k%2C+rho+%3A+tau+%3D+%28k+%2B+1%29%2F%28rho+-+2%29%3B+phi+%3D+k+*+%28rho+-+2%29+%2F+%28k+%2B+rho++-1%29


$$
k = (\tau + 1) \phi \\
\rho = \frac{\phi + 1}{\tau} + \phi + 2
$$

For $\tau, \phi > 0$. This parametrization doesn't allow you to have $\rho \leq 2$ but that is IMHO OK, because in this case the distribution has infinite variance (and for $\rho \leq 1$ the mean is not defined as well) - which seems mostly undesirable anyway.





##  Matching the mean and variance of the underlying beta prime distribution

https://en.wikipedia.org/wiki/Beta_prime_distribution
https://www.wolframalpha.com/input/?i=solve+for+r%2Ck%3A+x+%3D+r+%2F+%28k+-+1%29%3B+y+%3D+r%28r+%2B+k+-1%29+%2F%28%28k+-2%29+*+%28k+-+1%29%5E2%29

```{r}



tibble(mu = 50) %>% 
  crossing(tibble(y = seq(from = 0, to = 1000, by = 1))) %>% 
  crossing(tibble(mu_beta = c(0.1, 0.5, 1, 3, 5, 10)))%>% 
  crossing(tibble(var_beta = 10^seq(-2, 3))) %>% 
  mutate(
         rho = mu_beta * (mu_beta ^ 2 + mu_beta + var_beta) / var_beta,
         k = (mu_beta ^ 2 + mu_beta + var_beta) / var_beta,
         ld = waring_lpdf(y, mu, k = k , rho = rho),
         )  %>%
    ggplot(aes(x = y, y = ld, color = mu_beta, group = mu_beta)) + 
      geom_line() + 
      facet_wrap(~var_beta, ncol = 3, labeller = "label_both", scales = "free_y")



```

```{r}
main_data <- tibble(mu = 50) %>% 
  crossing(tibble(y = seq(from = 0, to = 1000, by = 1))) %>% 
  crossing(tibble(mu_beta = c(1, 5, 10)))%>% 
  crossing(tibble(var_beta = c(0.001, 0.1, 1))) %>% 
  mutate(
         rho = mu_beta * (mu_beta ^ 2 + mu_beta + var_beta) / var_beta,
         k = (mu_beta ^ 2 + mu_beta + var_beta) / var_beta,
         ld = waring_lpdf(y, mu, k = k , rho = rho),
         )  

nb_data <- main_data %>% filter(y %% 50 == 0) %>%
  mutate(
         variance = mu * (1 + (k + 1)/(rho - 2) + mu * (k + rho - 1) / ((rho-2) * k)),
         phi = mu ^ 2 / (variance - mu),
         nb_dens = dnbinom(y, mu = mu, size = phi, log = TRUE)
         )

main_data %>%
  ggplot(aes(x = y, y = ld, color = as.factor(mu_beta), group = mu_beta)) + 
      geom_point(data = nb_data, aes(y = nb_dens, shape = as.factor(mu_beta))) +
      geom_line() + 
      facet_wrap(~var_beta, ncol = 3, labeller = "label_both", scales = "free_y")
```




```{r}
tibble(mu = 50) %>% 
  crossing(tibble(y = seq(from = 0, to = 1000, by = 1))) %>% 
  crossing(tibble(mu_beta = c(0.01, 0.1, 0.3, 0.5, 0.7, 0.9,0.99)))%>% 
  crossing(tibble(prec_beta = 10^seq(-2, 3))) %>% 
  mutate(
      rho = mu_beta * prec_beta,
      k = prec_beta - rho,
      ld = waring_lpdf(y, mu, k = k, rho = rho))  %>%
    ggplot(aes(x = y, y = ld, color = mu_beta, group = mu_beta)) + geom_line() + facet_wrap(~prec_beta, ncol = 3, labeller = "label_both", scales = "free_y")
```
